---
title: "GOT1 knockdown vs control in pancreatic cancer — GSE157830"
subtitle: "GEO RNA-seq Portfolio · differential expression re-analysis"
author: "jam-sudo"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    fig-width: 7
    fig-height: 5
execute:
  warning: true
  message: false
---

```{r setup}
#| include: false
library(here)
source(here("scripts", "geo_helpers.R"))
suppressPackageStartupMessages({
  library(DESeq2)
  library(data.table)
  library(clusterProfiler)
})

ACC      <- "GSE157830"
PROJ_DIR <- here("projects", ACC)
DATA_DIR <- file.path(PROJ_DIR, "data")
FIG_DIR  <- file.path(PROJ_DIR, "results", "figures")
TAB_DIR  <- file.path(PROJ_DIR, "results", "tables")
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(TAB_DIR, recursive = TRUE, showWarnings = FALSE)

COUNTS_GZ  <- file.path(DATA_DIR, "raw", "GSE157830_genes.counts.txt.gz")
CLEAN_META <- file.path(DATA_DIR, "GSE157830_metadata_clean.csv")
```

## Biological question

**GOT1** (glutamic-oxaloacetic transaminase 1) is a cytosolic aspartate
aminotransferase that pancreatic ductal adenocarcinoma (PDAC) cells rely on to
maintain redox balance. This dataset (Kremer *et al.*, *"Labile Iron Release
Primes Pancreatic Cancer for Ferroptosis"*) uses a **doxycycline-inducible
shRNA** to knock down GOT1 in two PDAC cell lines and profiles the
transcriptome by RNA-seq.

> **Question:** Which genes and pathways change when GOT1 is knocked down in PDAC
> cells, consistently across two cell lines? We compare **GOT1 knockdown (+Dox)
> vs control (−Dox)** while statistically accounting for the cell-line baseline.

- **Accession:** [GSE157830](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE157830) · *Homo sapiens* · Illumina NovaSeq 6000 (GPL24676)
- **Design:** 2 cell lines (Tu8902, MiaPaCa2) × 2 treatments (control, GOT1 knockdown) × 3 replicates = 12 samples.
- A full pre-analysis suitability review is in [`notes/suitability.md`](notes/suitability.md).

## 1. Provenance: save verbatim GEO metadata

The cleaned metadata used for modelling is derived (and documented) in the
suitability note. Here we additionally pull the **verbatim** GEO sample metadata
for provenance. Network-dependent; wrapped so a failure does not break the report.

```{r geo-metadata}
meta_raw_path <- file.path(DATA_DIR, paste0(ACC, "_sample_metadata.csv"))
if (!file.exists(meta_raw_path)) {
  try({
    geo <- fetch_geo(ACC, dest_dir = DATA_DIR, get_supp = FALSE)
  }, silent = TRUE)
}
if (file.exists(meta_raw_path)) {
  cat("Verbatim GEO metadata saved at:", meta_raw_path, "\n")
} else {
  cat("GEO metadata fetch skipped/unavailable; cleaned metadata is used below.\n")
}
```

## 2. Load gene-level RSEM counts (rounded to integers)

The supplementary file `GSE157830_genes.counts.txt.gz` holds **RSEM expected
counts** with 8 annotation columns + 12 sample columns. RSEM counts are
fractional, so they are **rounded** before DESeq2 (the FPKM/TPM files are *not*
used). `ensembl_gene_id` is the unique row key; `gene_name` is kept as a label.

```{r load-counts}
raw <- data.table::fread(COUNTS_GZ)

anno_cols <- c("gene_name", "ensembl_gene_id", "gene_type", "entrez",
               "transcript_name", "ensembl_transcript", "transcript_type",
               "gene_description")
sample_cols <- setdiff(names(raw), anno_cols)
stopifnot(length(sample_cols) == 12)

gene_anno <- raw[, .(ensembl_gene_id, gene_name, entrez, gene_type)]
gene_anno[, gene_id := make.unique(as.character(ensembl_gene_id))]

counts <- as.matrix(raw[, ..sample_cols])
rownames(counts) <- gene_anno$gene_id
counts <- round(counts)                       # RSEM expected counts -> integers
storage.mode(counts) <- "integer"

cat("Count matrix:", nrow(counts), "genes x", ncol(counts), "samples\n")
cat("Fractional values were rounded (RSEM expected counts).\n")
```

## 3. Cleaned metadata & experimental design

Groups come from the GEO `genotype/variation` characteristics field, not the
free-text title (see suitability note, assumption A1). Reference levels:
`treatment = control`, `cell_line = Tu8902`.

```{r metadata}
meta <- read.csv(CLEAN_META, stringsAsFactors = FALSE)
rownames(meta) <- meta$sample_name
meta <- meta[sample_cols, ]                    # align to count columns
meta$cell_line <- factor(meta$cell_line, levels = c("Tu8902", "MiaPaCa2"))
meta$treatment <- factor(meta$treatment, levels = c("control", "GOT1_knockdown"))
meta$group     <- paste(meta$cell_line, meta$treatment, sep = " / ")

stopifnot(identical(colnames(counts), rownames(meta)))
write_results_table(meta, file.path(TAB_DIR, paste0(ACC, "_metadata_clean.csv")))
knitr::kable(meta[, c("geo_accession", "cell_line", "treatment", "dox", "replicate")],
             caption = "Cleaned sample metadata (design table)")
```

**Design formula:** `~ cell_line + treatment` — adjusts for the cell-line
baseline and estimates the shared GOT1-knockdown effect.
**Primary contrast:** `treatment` GOT1_knockdown vs control.

## 4. Differential expression (DESeq2)

```{r deseq2}
de <- run_deseq2(
  counts   = counts,
  coldata  = meta,
  design   = ~ cell_line + treatment,
  contrast = c("treatment", "GOT1_knockdown", "control")
)

# Shrunken LFCs for ranking/plots (apeglm).
res_shrink <- lfcShrink(de$dds, coef = "treatment_GOT1_knockdown_vs_control",
                        type = "apeglm")
de$res_df$log2FoldChange_shrunk <- res_shrink[de$res_df$gene_id, "log2FoldChange"]

# Attach in-file symbol + Entrez (more reliable than re-mapping).
anno_df <- as.data.frame(gene_anno)[, c("gene_id", "gene_name", "entrez")]
de$res_df <- dplyr::left_join(de$res_df, anno_df, by = "gene_id")
de$res_df <- dplyr::rename(de$res_df, symbol = gene_name)

write_results_table(de$res_df, file.path(TAB_DIR, paste0(ACC, "_DE_results.csv")))
summary(de$res)

# GOT1 itself — the knockdown target (positive control for the experiment).
knitr::kable(
  de$res_df[de$res_df$symbol == "GOT1" & !is.na(de$res_df$symbol),
            c("symbol", "baseMean", "log2FoldChange", "padj")],
  caption = "GOT1 (knockdown target) — expected strong down-regulation")
```

## 5. Quality control

```{r qc-basic}
grp <- meta$group
save_fig(plot_library_sizes(counts, grp),      file.path(FIG_DIR, "qc_library_sizes.png"))
save_fig(plot_count_distribution(counts, grp), file.path(FIG_DIR, "qc_count_distribution.png"),
         height = 6)
plot_library_sizes(counts, grp)
```

### PCA — colored by treatment, shaped by cell line

```{r qc-pca}
vsd_mat <- SummarizedExperiment::assay(de$vsd)
pc <- prcomp(t(vsd_mat))
ve <- round(100 * pc$sdev^2 / sum(pc$sdev^2), 1)
pca_df <- data.frame(PC1 = pc$x[, 1], PC2 = pc$x[, 2], meta)

p_pca <- ggplot(pca_df, aes(PC1, PC2, color = treatment, shape = cell_line)) +
  geom_point(size = 4, alpha = 0.9) +
  ggrepel::geom_text_repel(aes(label = replicate), size = 3, show.legend = FALSE) +
  scale_color_manual(values = c(control = "#1b9e77", GOT1_knockdown = "#d95f02")) +
  labs(title = "PCA (variance-stabilized counts)",
       x = paste0("PC1 (", ve[1], "%)"), y = paste0("PC2 (", ve[2], "%)"),
       color = "Treatment", shape = "Cell line") +
  geo_theme()
save_fig(p_pca, file.path(FIG_DIR, "qc_pca.png"))
p_pca
```

### Sample-sample correlation heatmap

```{r qc-corr}
plot_sample_correlation(vsd_mat, group = meta$treatment,
                        file = file.path(FIG_DIR, "qc_sample_correlation.png"))
plot_sample_correlation(vsd_mat, group = meta$treatment)
```

## 6. Differential-expression figures

```{r de-volcano}
save_fig(volcano_plot(de$res_df, title = "GOT1 knockdown vs control"),
         file.path(FIG_DIR, "volcano.png"))
volcano_plot(de$res_df, title = "GOT1 knockdown vs control")
```

```{r de-ma}
save_fig(ma_plot(de$res_df, title = "MA plot — GOT1 knockdown vs control"),
         file.path(FIG_DIR, "ma_plot.png"))
ma_plot(de$res_df, title = "MA plot — GOT1 knockdown vs control")
```

```{r de-heatmap}
top_gene_heatmap(de$vsd, de$res_df, n = 30, group = meta$treatment,
                 file = file.path(FIG_DIR, "top_genes_heatmap.png"),
                 id_col = "gene_id", label_col = "symbol")
```

```{r de-table}
top_tbl <- head(de$res_df[order(de$res_df$padj),
                          c("symbol", "gene_id", "baseMean",
                            "log2FoldChange", "pvalue", "padj")], 25)
knitr::kable(top_tbl, digits = 3, caption = "Top 25 differentially expressed genes")
```

## 7. Functional enrichment (GO biological process)

Significant genes (`padj < 0.05`, `|log2FC| > 1`) are tested for GO-BP
over-representation using the in-file **Entrez** IDs via clusterProfiler.

```{r enrichment}
sig <- subset(de$res_df, !is.na(padj) & padj < 0.05 &
                         abs(log2FoldChange) > 1 & !is.na(entrez))
cat("Significant genes for enrichment:", nrow(sig), "\n")

ego <- tryCatch(
  clusterProfiler::enrichGO(
    gene = unique(as.character(sig$entrez)),
    OrgDb = "org.Hs.eg.db", keyType = "ENTREZID",
    ont = "BP", pvalueCutoff = 0.05, qvalueCutoff = 0.1, readable = TRUE),
  error = function(e) { warning(conditionMessage(e)); NULL })

if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
  write_results_table(as.data.frame(ego),
                      file.path(TAB_DIR, paste0(ACC, "_GO_BP_enrichment.csv")))
  save_fig(barplot(ego, showCategory = 15) + ggtitle("GO BP enrichment"),
           file.path(FIG_DIR, "go_bp_enrichment.png"), height = 7)
  knitr::kable(head(as.data.frame(ego)[, c("ID", "Description", "p.adjust", "Count")], 15),
               caption = "Top GO biological-process terms")
} else {
  cat("No significant GO-BP enrichment returned.\n")
}
```

## 8. Interpretation

<!-- FILL after first render, grounded in the actual results tables/figures: -->

- **Positive control:** GOT1 should be strongly down-regulated in the knockdown
  arm (raw counts drop ~20–30×), confirming the perturbation worked.
- **Shared response:** genes consistently changed across both Tu8902 and
  MiaPaCa2 reflect the cell-line-independent consequence of losing GOT1.
- *Summarize the top up/down genes and the leading GO terms here, and relate
  them to GOT1's role in aspartate/redox metabolism and the paper's ferroptosis
  theme (iron handling, oxidative stress).*

## 9. Limitations

- **Control = uninduced (−Dox), not a non-targeting shRNA +Dox arm.** A
  doxycycline-intrinsic transcriptional effect cannot be fully separated from
  the GOT1-knockdown effect (suitability note A5).
- **RSEM expected counts were rounded** to satisfy DESeq2's integer requirement;
  importing per-transcript estimates via `tximport` would propagate uncertainty
  more rigorously.
- **Additive model only** (`~ cell_line + treatment`): it estimates a *shared*
  knockdown effect and does not test whether the response differs between cell
  lines (that would need `~ cell_line * treatment`).
- **Two cell lines, n = 3 each** — adequate for DE but a limited sampling of PDAC
  heterogeneity; findings are about these models, not PDAC in general.
- Annotation uses the submitter-provided gene/Entrez IDs; a small number of
  duplicated symbols/Ensembl IDs were de-duplicated (`make.unique`).

## Reproducibility

```{r session}
sessionInfo()
```
