---
title: "Luminal vs basal mammary epithelium across development — GSE60450"
subtitle: "GEO RNA-seq Portfolio · differential expression re-analysis"
author: "jam-sudo"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    fig-width: 7
    fig-height: 5
execute:
  warning: true
  message: false
---

```{r setup}
#| include: false
library(here)
source(here("scripts", "geo_helpers.R"))
suppressPackageStartupMessages({
  library(DESeq2)
  library(data.table)
  library(clusterProfiler)
})

ACC      <- "GSE60450"
ORGDB    <- "org.Mm.eg.db"
PROJ_DIR <- here("projects", ACC)
DATA_DIR <- file.path(PROJ_DIR, "data")
FIG_DIR  <- file.path(PROJ_DIR, "results", "figures")
TAB_DIR  <- file.path(PROJ_DIR, "results", "tables")
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(TAB_DIR, recursive = TRUE, showWarnings = FALSE)

COUNTS_GZ  <- file.path(DATA_DIR, "raw", "GSE60450_Lactation-GenewiseCounts.txt.gz")
CLEAN_META <- file.path(DATA_DIR, "GSE60450_metadata_clean.csv")
```

## Biological question

The mouse mammary gland is built from two epithelial lineages — **basal**
(myoepithelial) and **luminal** — and it remodels dramatically across
development (virgin → pregnancy → lactation). This dataset (Fu *et al.*, 2015)
FACS-sorted basal and luminal cells at three developmental stages and profiled
them by RNA-seq.

> **Question:** Which genes distinguish the **luminal** from the **basal**
> lineage, accounting for developmental stage?

- **Accession:** [GSE60450](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE60450) · *Mus musculus* · Illumina HiSeq 2000 (GPL13112)
- **Design:** 2 lineages (basal, luminal) × 3 stages (virgin, pregnant, lactating) × 2 mice = 12 samples.
- Suitability report: [`SUITABILITY.md`](SUITABILITY.md) (decision: **include**).

## 1. Load gene-wise counts (integer; no rounding needed)

`GSE60450_Lactation-GenewiseCounts.txt.gz` is an Rsubread `featureCounts`
matrix: `EntrezGeneID`, `Length`, then 12 sample columns. Values are already
integers. Row key = Entrez Gene ID.

```{r load-counts}
raw <- data.table::fread(COUNTS_GZ)
sample_cols <- setdiff(names(raw), c("EntrezGeneID", "Length"))
stopifnot(length(sample_cols) == 12)

counts <- as.matrix(raw[, ..sample_cols])
rownames(counts) <- as.character(raw$EntrezGeneID)
storage.mode(counts) <- "integer"

# Count columns are long codes like "MCL1-DG_BC2CTUACXX_..."; reduce to the code.
colnames(counts) <- sub("_.*$", "", colnames(counts))
cat("Count matrix:", nrow(counts), "genes x", ncol(counts), "samples\n")
cat("Sample codes:", paste(colnames(counts), collapse = ", "), "\n")
```

## 2. Cleaned metadata & design

Group labels were taken from the GEO `source_name_ch1` / sample titles, and the
count-column codes were mapped to samples via each GSM's **supplementary-file
name** (`GSM..._MCL1-XX_...`) — an authoritative key, not a guess. Reference
levels: `cell_type = basal`, `stage = virgin`.

```{r metadata}
meta <- read.csv(CLEAN_META, stringsAsFactors = FALSE)
rownames(meta) <- meta$code
meta <- meta[colnames(counts), ]                 # align to count columns by code
meta$cell_type <- factor(meta$cell_type, levels = c("basal", "luminal"))
meta$stage     <- factor(meta$stage, levels = c("virgin", "pregnant", "lactating"))
meta$group     <- paste(meta$cell_type, meta$stage, sep = " / ")

stopifnot(identical(colnames(counts), rownames(meta)))
write_results_table(meta, file.path(TAB_DIR, paste0(ACC, "_metadata_clean.csv")))
knitr::kable(meta[, c("geo_accession", "title", "cell_type", "stage", "replicate")],
             caption = "Cleaned design table (codes mapped via GSM supplementary filenames)")
```

**Design formula:** `~ stage + cell_type` — adjusts for developmental stage and
estimates the lineage (luminal vs basal) effect.
**Primary contrast:** `cell_type` luminal vs basal (n = 6 vs 6, well powered).

## 3. Differential expression (DESeq2)

```{r deseq2}
de <- run_deseq2(
  counts   = counts,
  coldata  = meta,
  design   = ~ stage + cell_type,
  contrast = c("cell_type", "luminal", "basal")
)
res_shrink <- lfcShrink(de$dds, coef = "cell_type_luminal_vs_basal", type = "apeglm")
de$res_df$log2FoldChange_shrunk <- res_shrink[de$res_df$gene_id, "log2FoldChange"]

de$res_df <- annotate_symbols(de$res_df, orgdb = ORGDB,
                              keytype = "ENTREZID", id_col = "gene_id")
write_results_table(de$res_df, file.path(TAB_DIR, paste0(ACC, "_DE_results.csv")))
summary(de$res)

# Lineage markers as a sanity check: Krt5/Krt14/Acta2 (basal) should be DOWN,
# Krt8/Krt18/Esr1 (luminal) UP, in luminal-vs-basal.
markers <- c("Krt5", "Krt14", "Acta2", "Krt8", "Krt18", "Esr1")
knitr::kable(de$res_df[match(markers, de$res_df$symbol),
                       c("symbol", "log2FoldChange", "padj")],
             caption = "Canonical lineage markers (sanity check)")
```

## 4. Quality control

```{r qc-basic}
grp <- meta$group
save_fig(plot_library_sizes(counts, grp),      file.path(FIG_DIR, "qc_library_sizes.png"))
save_fig(plot_count_distribution(counts, grp), file.path(FIG_DIR, "qc_count_distribution.png"), height = 6)
plot_library_sizes(counts, grp)
```

### PCA — colored by lineage, shaped by stage

```{r qc-pca}
vsd_mat <- SummarizedExperiment::assay(de$vsd)
pc <- prcomp(t(vsd_mat))
ve <- round(100 * pc$sdev^2 / sum(pc$sdev^2), 1)
pca_df <- data.frame(PC1 = pc$x[, 1], PC2 = pc$x[, 2], meta)

p_pca <- ggplot(pca_df, aes(PC1, PC2, color = cell_type, shape = stage)) +
  geom_point(size = 4, alpha = 0.9) +
  scale_color_manual(values = c(basal = "#1b9e77", luminal = "#d95f02")) +
  labs(title = "PCA (variance-stabilized counts)",
       x = paste0("PC1 (", ve[1], "%)"), y = paste0("PC2 (", ve[2], "%)"),
       color = "Lineage", shape = "Stage") +
  geo_theme()
save_fig(p_pca, file.path(FIG_DIR, "qc_pca.png"))
p_pca
```

### Sample-sample correlation

```{r qc-corr}
plot_sample_correlation(vsd_mat, group = meta$cell_type,
                        file = file.path(FIG_DIR, "qc_sample_correlation.png"))
plot_sample_correlation(vsd_mat, group = meta$cell_type)
```

## 5. DE figures

```{r de-figs}
save_fig(volcano_plot(de$res_df, title = "Luminal vs basal"),
         file.path(FIG_DIR, "volcano.png"))
save_fig(ma_plot(de$res_df, title = "MA plot — luminal vs basal"),
         file.path(FIG_DIR, "ma_plot.png"))
top_gene_heatmap(de$vsd, de$res_df, n = 30, group = meta$cell_type,
                 file = file.path(FIG_DIR, "top_genes_heatmap.png"),
                 id_col = "gene_id", label_col = "symbol")
volcano_plot(de$res_df, title = "Luminal vs basal")
```

```{r de-table}
knitr::kable(head(de$res_df[order(de$res_df$padj),
                            c("symbol", "gene_id", "baseMean", "log2FoldChange", "padj")], 25),
             digits = 3, caption = "Top 25 DE genes (luminal vs basal)")
```

## 6. Functional enrichment (GO BP)

```{r enrichment}
enr <- run_enrichment(de$res_df, orgdb = ORGDB, keytype = "ENTREZID",
                      id_col = "gene_id", padj_thresh = 0.05, lfc_thresh = 1)
if (!is.null(enr) && !is.null(enr$go_bp) && nrow(as.data.frame(enr$go_bp)) > 0) {
  write_results_table(as.data.frame(enr$go_bp),
                      file.path(TAB_DIR, paste0(ACC, "_GO_BP_enrichment.csv")))
  save_fig(barplot(enr$go_bp, showCategory = 15) + ggtitle("GO BP enrichment"),
           file.path(FIG_DIR, "go_bp_enrichment.png"), height = 7)
  knitr::kable(head(as.data.frame(enr$go_bp)[, c("ID", "Description", "p.adjust", "Count")], 15),
               caption = "Top GO biological-process terms")
} else cat("No significant GO-BP enrichment returned.\n")
```

## 7. Interpretation

*Grounded in the rendered results (18,418 genes tested; 7,347 significant at
`padj < 0.05`, `|log2FC| > 1`, adjusted for developmental stage).*

- **The lineage axis dominates — and the markers prove the labels are right.**
  Canonical **basal/myoepithelial** markers are strongly higher in basal cells
  (Krt5 log2FC ≈ −9.0, Acta2 ≈ −8.9, Krt14 ≈ −6.8) while **luminal** markers are
  higher in luminal cells (Krt18 ≈ +3.9, Esr1 ≈ +2.5, Krt8 ≈ +2.0). This is a
  clean biological positive control: the sorting, the metadata mapping, and the
  DESeq2 contrast all line up with known mammary biology.
- **Scale is expected, not alarming.** ~7,300 DE genes is large, but basal and
  luminal are genuinely distinct cell types — a deep transcriptional divide, not
  a subtle perturbation. (Contrast this with GSE157830's GOT1 knockdown, where a
  single-gene perturbation moved ~900 genes.)
- **Basal-high program:** GO biological process is led by **extracellular-matrix
  organization, cell-substrate adhesion, and muscle contraction** — exactly the
  contractile myoepithelial phenotype of basal cells.
- **Luminal-high program:** top luminal genes include **Tph1** (serotonin
  biosynthesis, a known luminal/lactation gene), **Calca**, and secretory
  **Wfdc** family members — the secretory luminal identity.

## 8. Limitations

- **Only 2 biological replicates per group** (2 mice × 6 groups). The luminal-vs-
  basal contrast pools to n = 6 per side and is well powered, but stage-specific
  contrasts (n = 2) are low-powered; treat those as exploratory.
- Additive design `~ stage + cell_type` assumes the lineage difference is
  consistent across stages; a `stage:cell_type` interaction would test lineage-
  specific developmental remodelling.
- Gene identifiers are Entrez IDs from the submitter's count matrix; a few may be
  retired/merged in current annotation.

## Reproducibility

```{r session}
sessionInfo()
```
