---
title: "RNA-seq Differential Expression — GSEXXXXXX"
subtitle: "GEO RNA-seq Portfolio"
author: "Your Name"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    fig-width: 7
    fig-height: 5
params:
  accession: "GSEXXXXXX"
  organism_db: "ORGANISM_DB"   # e.g. org.Hs.eg.db (human), org.Mm.eg.db (mouse)
  keytype: "ENSEMBL"            # ID type of the count rownames: ENSEMBL / SYMBOL / ENTREZID
  group_col: null              # metadata column defining the comparison (set after inspecting samples)
  contrast: null               # c(group_col, numerator, denominator), e.g. c("condition","KO","WT")
execute:
  warning: true
  message: false
---

```{r setup}
#| include: false
source(here::here("scripts", "geo_helpers.R"))
library(here)

ACC      <- params$accession
DATA_DIR <- here("projects", ACC, "data")
FIG_DIR  <- here("projects", ACC, "results", "figures")
TAB_DIR  <- here("projects", ACC, "results", "tables")
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(TAB_DIR, recursive = TRUE, showWarnings = FALSE)
```

## Biological question

<!-- FILL: State the biological question in 2-3 sentences. What conditions are
     compared, in what tissue/cell type, and what would a meaningful result mean? -->

> *Placeholder — replace with the dataset's biological question.*

**GEO accession:** `{r} ACC`
**Organism annotation DB:** `{r} params$organism_db`

## 1. Data acquisition

We download the GEO series matrix (sample metadata + any matrix expression
values) and the supplementary files (raw counts, when the authors deposited
them there). The raw sample metadata is written to disk unmodified.

```{r acquire}
geo <- fetch_geo(ACC, dest_dir = DATA_DIR, get_supp = TRUE)
cat("Samples:", nrow(geo$pheno), " | Supplementary files:", length(geo$supp_files), "\n")
```

## 2. Data-type assessment (decision gate)

Before any differential expression, we determine **what GEO actually provided**.
Raw counts → DESeq2 is valid. Normalized values only → DESeq2 is *not* valid and
we say so explicitly. Microarray / unusable → we flag the dataset.

```{r assess}
verdict <- assess_data_type(geo)
print(verdict)

# Flags used to conditionally evaluate later chunks.
can_deseq2 <- isTRUE(verdict$suitable_for_deseq2)
```

```{r assess-flag}
#| echo: false
#| results: asis
if (!can_deseq2) {
  cat("::: {.callout-warning}\n")
  cat("## This dataset is not standard raw-count RNA-seq input\n\n")
  cat(verdict$reason, "\n\n**Recommendation:** ", verdict$recommendation, "\n")
  cat(":::\n")
}
```

<!-- If can_deseq2 is FALSE: load the count matrix below is skipped. Document the
     limitation in the Limitations section and, if appropriate, run a limma-based
     analysis on the normalized values instead (clearly labelled as such). -->

## 3. Load count matrix

<!-- FILL: Adjust this chunk to read the actual count file for THIS dataset.
     Supplementary file names vary. `verdict$count_files` lists candidates. -->

```{r load-counts}
#| eval: !expr can_deseq2
# Example for a typical tab-delimited gene x sample count file:
# counts_path <- verdict$count_files[1]
# counts <- as.matrix(data.table::fread(counts_path), rownames = 1)
#
# Placeholder so the template renders before customization:
stopifnot("Set the count-loading code for this dataset." = exists("counts"))
```

## 4. Sample metadata & experimental design

```{r samples}
samp <- summarize_samples(geo$pheno)
cat("Candidate grouping columns:\n"); print(samp$candidate_group_cols)
knitr::kable(samp$preview, caption = "Sample metadata preview")

# Save a tidy metadata summary for the portfolio.
write_results_table(geo$pheno, file.path(TAB_DIR, paste0(ACC, "_sample_metadata.csv")))
```

```{r contrasts}
#| eval: !expr (!is.null(params$group_col))
proposed <- propose_contrasts(geo$pheno, params$group_col)
cat("Group sizes:\n"); print(proposed$group_sizes)
cat("\nProposed pairwise contrasts:\n"); str(proposed$contrasts, max.level = 2)
```

**Design formula:** `~ {r} if (!is.null(params$group_col)) params$group_col else "<group_col>"`
(make the design explicit; add batch/covariate terms here if the metadata supports it).

## 5. Quality control

```{r qc}
#| eval: !expr can_deseq2
group <- if (!is.null(params$group_col)) factor(geo$pheno[[params$group_col]]) else NULL

save_fig(plot_library_sizes(counts, group),      file.path(FIG_DIR, "qc_library_sizes.png"))
save_fig(plot_count_distribution(counts, group), file.path(FIG_DIR, "qc_count_distribution.png"))

# PCA / sample correlation use a variance-stabilized matrix once DESeq2 has run.
```

## 6. Differential expression

```{r de}
#| eval: !expr (can_deseq2 && !is.null(params$group_col) && !is.null(params$contrast))
coldata <- geo$pheno[colnames(counts), , drop = FALSE]
coldata[[params$group_col]] <- factor(coldata[[params$group_col]])

de <- run_deseq2(
  counts   = counts,
  coldata  = coldata,
  design   = as.formula(paste("~", params$group_col)),
  contrast = params$contrast
)

# QC that needs the vst object.
save_fig(plot_pca(SummarizedExperiment::assay(de$vsd),
                  coldata[[params$group_col]], "PCA (variance-stabilized)"),
         file.path(FIG_DIR, "qc_pca.png"))
plot_sample_correlation(SummarizedExperiment::assay(de$vsd),
                        coldata[[params$group_col]],
                        file = file.path(FIG_DIR, "qc_sample_correlation.png"))

# Annotate symbols and save the full results table.
de$res_df <- annotate_symbols(de$res_df, orgdb = params$organism_db,
                              keytype = params$keytype)
write_results_table(de$res_df, file.path(TAB_DIR, paste0(ACC, "_DE_results.csv")))

summary(de$res)
```

```{r de-figs}
#| eval: !expr (can_deseq2 && !is.null(params$group_col) && !is.null(params$contrast))
save_fig(volcano_plot(de$res_df), file.path(FIG_DIR, "volcano.png"))
save_fig(ma_plot(de$res_df),      file.path(FIG_DIR, "ma_plot.png"))
top_gene_heatmap(de$vsd, de$res_df, n = 30,
                 group = coldata[[params$group_col]],
                 file = file.path(FIG_DIR, "top_genes_heatmap.png"))

knitr::kable(head(de$res_df, 20), caption = "Top 20 differentially expressed genes")
```

## 7. Functional enrichment

```{r enrich}
#| eval: !expr (can_deseq2 && exists("de"))
enr <- run_enrichment(de$res_df, orgdb = params$organism_db, keytype = params$keytype)
if (!is.null(enr) && !is.null(enr$go_bp) && nrow(as.data.frame(enr$go_bp)) > 0) {
  write_results_table(as.data.frame(enr$go_bp),
                      file.path(TAB_DIR, paste0(ACC, "_GO_BP_enrichment.csv")))
  knitr::kable(head(as.data.frame(enr$go_bp)[, c("ID","Description","p.adjust","Count")], 15),
               caption = "Top GO biological-process terms")
} else {
  cat("No significant GO enrichment (or annotation unavailable).\n")
}
```

## 8. Interpretation

<!-- FILL: 1-2 paragraphs. What are the top genes/pathways? Do they fit the
     known biology of the comparison? What is the headline takeaway? Write for
     both a technical reviewer and an interested non-specialist. -->

> *Placeholder — replace with the biological interpretation of the results.*

## 9. Limitations

<!-- FILL honestly. Examples to consider:
     - Data type (raw counts vs normalized-only) and what that constrains.
     - Sample size / unbalanced groups / no replicates.
     - Confounders or batch effects in the metadata.
     - Annotation gaps (ID mapping coverage).
     - Whether the chosen contrast is the only valid one. -->

- *Placeholder — list dataset-specific limitations here.*
- Data-type verdict for this dataset: **`{r} verdict$type`** — `{r} verdict$reason`

## Reproducibility

```{r session}
sessionInfo()
```
