---
title: "IgG-positive vs IgG-negative pancreatic cancer cells — GSE334534"
subtitle: "GEO RNA-seq Portfolio · differential expression re-analysis"
author: "jam-sudo"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    fig-width: 7
    fig-height: 5
execute:
  warning: true
  message: false
---

```{r setup}
#| include: false
library(here)
source(here("scripts", "geo_helpers.R"))
suppressPackageStartupMessages({
  library(DESeq2)
  library(data.table)
  library(clusterProfiler)
})

ACC      <- "GSE334534"
ORGDB    <- "org.Hs.eg.db"
PROJ_DIR <- here("projects", ACC)
DATA_DIR <- file.path(PROJ_DIR, "data")
FIG_DIR  <- file.path(PROJ_DIR, "results", "figures")
TAB_DIR  <- file.path(PROJ_DIR, "results", "tables")
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(TAB_DIR, recursive = TRUE, showWarnings = FALSE)

COUNTS_GZ  <- file.path(DATA_DIR, "raw", "GSE334534_genes_expected_counts_all_samples.txt.gz")
CLEAN_META <- file.path(DATA_DIR, "GSE334534_metadata_clean.csv")
```

## Biological question

This dataset profiles **PL45 pancreatic-cancer cells** that were sorted by
surface **IgG binding** (IgG-positive vs IgG-negative), across two single-cell
clones (A1, A14), to find transcriptional correlates of humoral-immunity
phenotypes.

> **Question:** Which genes differ between **IgG-positive** and **IgG-negative**
> PL45 cells, consistently across the two clones?

- **Accession:** [GSE334534](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE334534) · *Homo sapiens* · bulk RNA-seq.
- **Design:** 2 clones (A1, A14) × 2 sorts (IgG_Negative, IgG_Positive) × 3 replicates = 12 samples.

> **Note (why this isn't the template):** GSE334534 has **no GEO series-matrix
> file** (`matrix/` returns 404), so the generic `getGEO()`-based acquisition in
> the template fails. This curated report instead reads the **local count file**
> and a cleaned metadata table built from the GEO `characteristics` fields.

## 1. Load RSEM expected counts (rounded to integers)

`GSE334534_genes_expected_counts_all_samples.txt.gz` holds **RSEM expected
counts** (`gene_id` = HGNC symbol + 12 sample columns). RSEM counts are
fractional, so they are **rounded** before DESeq2.

```{r load-counts}
raw <- data.table::fread(COUNTS_GZ)
sample_cols <- setdiff(names(raw), "gene_id")
gene_ids <- make.unique(as.character(raw$gene_id))
counts <- as.matrix(raw[, ..sample_cols])
rownames(counts) <- gene_ids
colnames(counts) <- sub("^expected_count_", "", colnames(counts))  # strip prefix
counts <- round(counts)                                            # RSEM -> integers
storage.mode(counts) <- "integer"
cat("Count matrix:", nrow(counts), "genes x", ncol(counts), "samples\n")
```

## 2. Cleaned metadata & design

Group labels come from the GEO `characteristics` fields (`treatment`, `batch`),
saved verbatim per sample; the **clone** (A1/A14) is parsed from the sample name
(it is *not* a separate characteristics field — `cell line` is constant `PL45`).

```{r metadata}
meta <- read.csv(CLEAN_META, stringsAsFactors = FALSE)
rownames(meta) <- meta$sample
meta <- meta[colnames(counts), ]                       # align to count columns
meta$clone     <- factor(meta$clone, levels = c("A1", "A14"))
meta$treatment <- factor(meta$treatment, levels = c("IgG_Negative", "IgG_Positive"))
meta$batch     <- factor(meta$batch)
meta$group     <- paste(meta$clone, meta$treatment, sep = " / ")
stopifnot(identical(colnames(counts), rownames(meta)))
write_results_table(meta, file.path(TAB_DIR, paste0(ACC, "_metadata_clean.csv")))
knitr::kable(meta[, c("geo_accession", "clone", "treatment", "batch", "replicate")],
             caption = "Cleaned design table")
```

**Design:** `~ clone + treatment`. The clone factor is balanced with treatment
(each clone has 3 IgG_Negative + 3 IgG_Positive). **Batch is deliberately not
included**: it is partially confounded with treatment (IgG_Negative is mostly
batch 1, IgG_Positive mostly batch 3), so adding it would absorb the treatment
signal (see Limitations).

```{r batch-confounding}
knitr::kable(table(meta$treatment, meta$batch), caption = "treatment × batch (note the imbalance)")
```

## 3. Differential expression (DESeq2)

```{r deseq2}
de <- run_deseq2(
  counts   = counts,
  coldata  = meta,
  design   = ~ clone + treatment,
  contrast = c("treatment", "IgG_Positive", "IgG_Negative")
)
res_shrink <- lfcShrink(de$dds, coef = "treatment_IgG_Positive_vs_IgG_Negative", type = "apeglm")
de$res_df$log2FoldChange_shrunk <- res_shrink[de$res_df$gene_id, "log2FoldChange"]

# gene_id is already the HGNC symbol.
de$res_df$symbol <- de$res_df$gene_id
write_results_table(de$res_df, file.path(TAB_DIR, paste0(ACC, "_DE_results.csv")))
summary(de$res)
cat("Significant (padj<0.05, |log2FC|>1):",
    sum(!is.na(de$res_df$padj) & de$res_df$padj < 0.05 & abs(de$res_df$log2FoldChange) > 1), "\n")
```

## 4. Quality control

```{r qc}
grp <- meta$group
save_fig(plot_library_sizes(counts, grp),      file.path(FIG_DIR, "qc_library_sizes.png"))
save_fig(plot_count_distribution(counts, grp), file.path(FIG_DIR, "qc_count_distribution.png"), height = 6)

vsd_mat <- SummarizedExperiment::assay(de$vsd)
pc <- prcomp(t(vsd_mat)); ve <- round(100 * pc$sdev^2 / sum(pc$sdev^2), 1)
pca_df <- data.frame(PC1 = pc$x[,1], PC2 = pc$x[,2], meta)
p_pca <- ggplot(pca_df, aes(PC1, PC2, color = treatment, shape = clone)) +
  geom_point(size = 4, alpha = 0.9) +
  scale_color_manual(values = c(IgG_Negative = "#1b9e77", IgG_Positive = "#d95f02")) +
  labs(title = "PCA (variance-stabilized)", color = "Sort", shape = "Clone",
       x = paste0("PC1 (", ve[1], "%)"), y = paste0("PC2 (", ve[2], "%)")) + geo_theme()
save_fig(p_pca, file.path(FIG_DIR, "qc_pca.png"))
plot_sample_correlation(vsd_mat, group = meta$treatment,
                        file = file.path(FIG_DIR, "qc_sample_correlation.png"))
p_pca
```

## 5. DE figures

```{r de-figs}
save_fig(volcano_plot(de$res_df, title = "IgG-positive vs IgG-negative"), file.path(FIG_DIR, "volcano.png"))
save_fig(ma_plot(de$res_df, title = "MA plot — IgG-positive vs IgG-negative"), file.path(FIG_DIR, "ma_plot.png"))
top_gene_heatmap(de$vsd, de$res_df, n = 30, group = meta$treatment,
                 file = file.path(FIG_DIR, "top_genes_heatmap.png"), id_col = "gene_id", label_col = "symbol")
volcano_plot(de$res_df, title = "IgG-positive vs IgG-negative")
```

```{r de-table}
knitr::kable(head(de$res_df[order(de$res_df$padj),
                            c("symbol", "baseMean", "log2FoldChange", "padj")], 25),
             digits = 3, caption = "Top 25 DE genes")
```

## 6. Functional enrichment (GO BP)

```{r enrich}
enr <- run_enrichment(de$res_df, orgdb = ORGDB, keytype = "SYMBOL",
                      id_col = "gene_id", padj_thresh = 0.05, lfc_thresh = 1)
if (!is.null(enr) && !is.null(enr$go_bp) && nrow(as.data.frame(enr$go_bp)) > 0) {
  write_results_table(as.data.frame(enr$go_bp), file.path(TAB_DIR, paste0(ACC, "_GO_BP_enrichment.csv")))
  save_fig(barplot(enr$go_bp, showCategory = 15) + ggtitle("GO BP enrichment"),
           file.path(FIG_DIR, "go_bp_enrichment.png"), height = 7)
  knitr::kable(head(as.data.frame(enr$go_bp)[, c("ID","Description","p.adjust","Count")], 15),
               caption = "Top GO biological-process terms")
} else cat("No significant GO-BP enrichment returned.\n")
```

## 7. Interpretation

*Grounded in the rendered results (15,434 genes tested; **437 significant** at
`padj < 0.05`, `|log2FC| > 1`, clone-adjusted).*

- **A modest, real difference** separates IgG-positive from IgG-negative PL45
  cells. Genes higher in **IgG-positive** cells include epithelial/differentiation
  factors — **GRHL2** (log2FC ≈ +1.9), **SORL1**, **GALNT5**, **TNS4**,
  **SH3TC2** — hinting at an epithelial-program shift in the IgG-binding fraction.
- **Read with caution — the signal is noisy.** Several of the most extreme
  features are rRNA/snRNA / pseudogenes (RNA5-8S5, RNU5F-1, NUDT4P2), and the GO
  enrichment is dominated by **taste-receptor** and **snRNP-assembly** terms —
  patterns that usually reflect technical/non-coding artifacts rather than a
  clean biological program.
- **Confounding matters here.** Batch is partially confounded with the IgG sort
  (Limitations), so part of this modest effect could be technical. A
  batch-balanced re-design, or filtering rRNA/snRNA features, would be the right
  way to sharpen the biological signal before drawing strong conclusions.

> Honest bottom line: GSE334534 is *suitable* for DESeq2 (raw counts, balanced
> clone×sort design) and yields a real but **modest, partly-confounded** result —
> a good example of a dataset that passes triage yet warrants caveats, not a
> headline finding.

## 8. Limitations

- **Batch is partially confounded with the IgG sort** (IgG_Negative ≈ batch 1,
  IgG_Positive ≈ batch 3), so a batch-corrected model is not identifiable; the
  reported effect may include a residual batch component. The clone-adjusted
  design is the cleanest identifiable model here.
- **RSEM expected counts were rounded** to integers for DESeq2.
- **Clone (A1/A14) parsed from the sample name**, not a `characteristics` field
  (`cell line` is constant `PL45`); confirmed by consistent name tokens.
- Two clones, n = 3 per clone×sort — adequate for DE but a limited sampling.
- GSE334534 has no GEO series matrix; counts are the submitter's RSEM table.

## Reproducibility

```{r session}
sessionInfo()
```
